Shifts in water supply and demand drive land cover change across Chile

### 2025-04-08 article/article_scientific_reports/mansucript_11012025.tex
actualización paper

Expanding analyses to include multiple dimensions of droughts can
provide complementary insights into the Earth's water balance - and its
impacts - over multiple time scales. Yet, the World Meteorological
Organization recommends the use of a single drought index for monitoring
droughts \citep{WMO2012}, i.e., the multi-scale Standardized
Precipitation Index (SPI; \citep{McKee1993}), which is limited in that
it only considers water supply in the form of precipitation. The
Standardized Precipitation Evapotranspiration Index (SPEI;
\citep{Vicente-Serrano2020}) builds upon SPI by incorporating the
effects of temperature on drought, and is now used widely for drought
monitoring \citep[e.g.,][]{Gebrechorkos2023, Liu2024}. Indices derived
from soil moisture products \citep{Narasimhan2005, Souza2021}, such as
the Standardized Soil Moisture Index
\citep[SSI,][]{AghaKouchak2014, AghaKouchak2015} also monitor water
supply and are thought to better capture water availability for crops,
thus providing more relevant information for evaluating agricultural
droughts. To disentangle the effects of precipitation from those of
temperature \citep{Vicente-Serrano2020}, as well as to capture droughts
in terms of water atmospheric demand, AED has been integrated into the
Evaporative Demand Drought Index \citep[EDDI,][]{McEvoy2016}, which is
particularly effective at detecting the rapid onset or intensification
of droughts. To quantify vegetation water demand, we must use the actual
evapotranspiration, or the amount of water removed from a surface by
evaporation and transpiration; the Standardized Evapotranspiration Index
(SETI; \citep{Yang2019}) can be used for this purpose. In turn,
ecological droughts, which capture the joint effects of precipitation
and temperature in modifying natural and productive ecosystems
\citep{Camps-Valls2021, Paruelo2016, Helman2014}, are complex to measure
and can therefore be monitored using multiple drought indices that
capture the multiple dimensions of drought, e.g., precipitation,
temperature, evapotranspiration, and AED. Although such an approach
accounts for the joint effects of changes in natural and productive
ecosystems, its potential impacts on land cover change have been largely
unexplored.

@@

To derive the drought indices of water supply and demand we used the
ERA5L dataset and the MODIS product (MOD13A3 \citep{Didan2015b}), with a
monthly frequency for 2000--2023. Drought indices capture historical
anomalies of water supply and demand. To quantify each anomaly, the
common practice is to derive it following a statistical parametric
method in which it is assumed that the statistical distribution of the
data is known \citep{Heim2002}. The use of an erroneous statistical
distribution that does not fit the data is usually the highest source of
uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
high degree of climatic variability, it is difficult to choose a
statistical distribution that can be used across its entire extent. We
therefore use a non-parametric method for the calculation of the drought
indices, following \citep{Farahmand2015}.

@@

We also used the MODIS product to calculate vegetation productivity, and
calculated anomalies of cumulative NDVI using zcNDVI
\citep{Zambrano2018}, which was derived from the monthly time series of
NDVI, with Equations 2 and 4. For vegetation productivity, we selected
the time scale that best correlates with annual net primary productivity
(NPP) across continental Chile. For this purpose, we calculated zcNDVI
for time scales of 1, 3, 6, and 12 months (from December) and compared
it with the annual NPP. We obtained NPP from MOD17A3HGF90. We chose to
use six months because the \(R^2\) between zcNDVI and NPP reaches its
highest value at six months, obtaining an \(R^2\) of 0.31 for forest and
0.72 for shrubland (Supplementary Information Section S2). We
subsequently used zcNDVI with a time scale of 6 months and calculated it
at a monthly frequency for 2000--2023.

@@

We discovered that the Atacama desert, Central Andean dry puna, and the
Magellanic subpolar forests experience an increase in water supply (SPI,
SSI), as well as an increase in atmospheric and vegetation water demand
(EDDI, SETI). However, in the Magellanic subpolar forests, there is no
significant increase or decrease in SETI across time scales. Also, we
found a significant decreasing trend in water supply (SPI, SPEI, and
SSI) across the Southern Andean steppe, Chilean Matorral
\citep{Boisier2018, Sarricolea2019}, Valdivian temperate forests, and
Patagonian steppe, accompanied by an increase in atmospheric water
demand (EDDI). Our results indicate that the trend of water supply and
atmospheric demand tends to decrease or increase more strongly over
longer time scales---a trend that is consistent with the progressive
intensification of drought severity across much of Chile, as observed in
other regions facing long-term droughts \citep{Rashid2019, Miro2023}.
Simultaneously, we observed a divergent trend between the EDDI and SETI
in various ecoregions. In the majority of ecoregions, a rise in
atmospheric water demand (EDDI) typically leads to a rise in vegetation
water demand (SETI). However, in the zones most affected by drought
(Fig. 3 and Fig. 5), where water availability is limited, an increase in
atmospheric water demand results in a decrease in vegetation water
demand. The Chilean matorral and the Patagonian steppe are where this is
most evident. When combined, our findings demonstrate a persistent
drying trend in the Chilean matorral, the Patagonian steppe, and the
Southern Andean steppe. We attribute this trend to a simultaneous
decrease in precipitation and an increase in atmospheric evaporative
demand, which leads to a decrease in the water demand for vegetation in
water-limited zones\citep{Pascoa2021}.

@@

The consequences of the persistent drying trend for ecosystems
throughout continental Chile are manifold. First, the prolonged
hydrological drought, i.e., precipitation deficit, has reduced
groundwater storage (SSI, \citep{Taucare2024}), leading to a steady
decline in vegetation productivity (zcNDVI) since 2000 across the
Chilean Matorral, Patagonian steppe, and Southern Andean steppe,
reaching its lowest level between 2020 and 2022, which could be due to
either a decrease in vegetation area, a loss of biomass or browning in
forest ecosystems. Reports for natural and productive ecosystems
\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021} most strongly associate
this decline with declines in soil moisture and increases in
evapotranspiration. Second, the sharp decline in vegetation productivity
in the Chilean Matorral and Valdivian temperate forests showed that
grasses and shrublands respond to shifts in water supply on higher time
scales (12 months), in comparison to savanna and croplands (6 months).
Also, in the Valdivian temperate forest that has a higher surface with
forest trees, soil moisture (SSI) and vegetation water demand (SETI) at
higher time scales, 12 and 36 months, respectively, evidence stronger
correlations. Which is consistent with recent studies showing that
progressive, long-term water deficits in central Chile have triggered
forest browning and declines in native forest productivity
\citep[e.g.,][]{Miranda2023, Miranda2020, Venegas2022}. While our
analysis do not distinguish between native and planted forests, the
latter of which are considered to be more drought tolerant in central
and southern Chile \citep{Carrasco2022}, we show that forest area
declines more sharply in response to increasing water demand due to
rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI;
refs. \citep{Fajardo2019, Holz2018}), which may have cascading impacts
on multiple facets of forest diversity \citep{Segovia2021, Sabatini2022}

Moreover, the strengthening of the correlation between vegetation
productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI)
over multiple time scales (up to 36 months) and across land cover types
(Fig. 5) - demonstrates the impacts of climate change on the water
balance in Chile. Impacts may extend beyond vegetation productivity, as
reduced soil moisture in central Chile and the western United States has
increased wildfire activity\citep{Holden2018, Gonzalez2018}, which is a
growing concern in Chile and may be further exacerbated by extensive
plantations of highly flammable tree species, e.g., Eucalyptus spp. and
Pinus spp. \citep{Bowman2019}. Third, we found that the decline in the
vegetation productivity of croplands is due to a decrease in the water
supply and vegetation water demand to a greater extent than to an
increase in atmospheric water demand \citep{Quiring2010}, causing a
decline in water availability. This is consistent with evidence that
more water-intensive crops have replaced less water-intensive crops in
central Chile, leading to an increase in water extraction from rivers or
groundwater \citep{Munoz2020, Duran2020}.

@@

We found evidence that temporal decreases in water supply (SPEI, SSI)
and decreases in vegetation water demand (SETI) are driving shifts not
only in vegetation productivity but also in land cover across most of
continental Chile. Shrubland covers were the land cover type most
sensitive to changes in the water balance over short and long temporal
scales. In contrast, forest surfaces are the type less affected by
drought. Moreover, our results provide evidence that, in addition to
shrublands cover, other land cover types have been affected by water
deficits, particularly savannas and grasslands. Our results therefore
suggest that multiple land cover types could be vulnerable to regime
shifts towards more drought tolerant land cover types
\citep{Scheffer2001, Martinez-Vilalta2016}, such as forest, whose cover
increased non-linearly in response to increasing drought intensity.
Changes in cropland cover are not a direct consequence of drought (Fig.
7), but rather an indirect one, likely occupying burned areas and
possibly reflecting the decision of resource-poor farmers to migrate to
regions with more abundant water resources or to change economic
activity \citep{AghaKouchak2021, Hermans2021}. In contrast, the increase
in shrubland cover due to a decrease in savanna cover may be ecological,
as shrubs may be more drought tolerant than other growth
forms\citep{Eldridge2011, Gotmark2016}.

We run each model ten times to account for the uncertainty in its
performance. In the supplementary material (Fig. S12-S17), we show the
error bar of r-squared; despite the inherent uncertainty, the analysis
reinforces our results regarding how drought impacts land cover type.

Overall, our results show that long-term declines in water supply and
demand have induced widespread, multi-dimensional impacts on the
vegetation productivity and on the extent of land cover types. While
prolonged droughts may directly cause shifts to more drought-tolerant
land cover types, such as shrublands, they may also influence land cover
change through human decision making and activities. This study extends
current understanding of drought impacts by demonstrating how their
multidimensionality emerges over multiple time scales and across land
cover types, which can contribute to developing context-specific
adaptation strategies for agriculture, biodiversity conservation, and
natural resource management.

@@

This study has evaluated the impact of water supply and demand on
vegetation productivity and how land use change influences it, with some
important limitations to highlight. One of the major limitations of the
study is the consideration of secondary information. For instance, we
used estimates of water supply and demand, such as ERA5L and MODIS,
which, despite their improved estimation capacity, suffer from biases
and uncertainties \citep{Gomis2023, Clelland2024} in different areas or
climatic conditions. In this study, we compared the ERA5L data with
climatic stations (see Table S2) to verify bias and uncertainty, but
future studies will need to correct the products for a more precise
measurement. We used zcNDVI \citep{Zambrano2018} (MODIS) as a proxy for
vegetation productivity, which has proven to be a good estimate of NPP
(see Fig. S1 and S2), but its quality varies between different types of
vegetation.

To define land cover, products were used that estimate the classes using
classification models, which are subject to quality errors that must be
taken into account \citep{Stehman2019, Verburg2011}. In addition, in our
case we used macro classes of land cover, where, for example, the
different types of forests were grouped (e.g., monoculture, native
forest). This approach hinders our ability to understand the effects of
drought on the various subclasses within each land use class. When it
comes to cultivated areas, we cannot distinguish between rainfed and
irrigated areas using macro classes. However, in this study, we aimed to
provide a broad overview at a national scale, with the goal of focusing
on a more detailed level in future research.

Other studies have analyzed how human decisions impact land use change,
such as population increase \citep{Kleemann2017}. In our case, we
attempted to consider this by incorporating road density, night lights,
and burned areas. However, human decisions are complex to consider from
the point of view of earth observation. On the other hand, earth
observation tools can analyze land cover change, but each type's use
depends on social and economic factors, which are challenging to
quantify \citep{Rindfuss2004, Jenner2024} and necessitate the
integration of social, natural, and geographic information sciences.

>>>>

Expanding analyses to include multiple dimensions of droughts can
provide complementary insights into the Earth's water balance - and its
impacts - over multiple time scales. Yet, the World Meteorological
Organization recommends the use of a single drought index for monitoring
droughts \citep{WMO2012}, i.e., the multi-scale Standardized
Precipitation Index (SPI; \citep{McKee1993}), which is limited in that
it only considers water supply in the form of precipitation. The
Standardized Precipitation Evapotranspiration Index (SPEI;
\citep{Vicente-Serrano2020}) builds upon SPI by incorporating the
effects of temperature on drought, and is now used widely for drought
monitoring \citep[e.g.,][]{Gebrechorkos2023, Liu2024}. Indices derived
from soil moisture products \citep{Narasimhan2005, Souza2021}, such as
the Standardized Soil Moisture Index
\citep[SSI,][]{AghaKouchak2014, AghaKouchak2015} also monitor water
supply and are thought to better capture water availability for crops,
thus providing more relevant information for evaluating agricultural
droughts. To disentangle the effects of precipitation from those of
temperature \citep{Vicente-Serrano2020}, as well as to capture droughts
in terms of water atmospheric demand, AED has been integrated into the
Evaporative Demand Drought Index \citep[EDDI,][]{McEvoy2016}, which is
particularly effective at detecting the rapid onset or intensification
of droughts. To quantify vegetation water demand, we must use the actual
evapotranspiration, or the amount of water removed from a surface by
evaporation and transpiration; the Standardized Evapotranspiration Index
(SETI; \citep{Yang2019}) can be used for this purpose. In turn,
ecological droughts, which capture the joint effects of precipitation
and temperature in modifying natural and productive ecosystems
\citep{Camps-Valls2021, Paruelo2016, Helman2014}, are complex to measure
and can therefore be monitored using multiple drought indices that
capture the multiple dimensions of drought, e.g., precipitation,
temperature, evapotranspiration, and AED. Although such an approach
accounts for the joint effects of changes in natural and productive
ecosystems, its potential impacts on land cover change have been largely
unexplored \citep{Tollerud2018, Marumbwa2020}.

@@

To derive the drought indices of water supply and demand we used the
ERA5L dataset, with a monthly frequency for 2000--2023. Drought indices
capture historical anomalies of water supply and demand. To quantify
each anomaly, the common practice is to derive it following a
statistical parametric method in which it is assumed that the
statistical distribution of the data is known \citep{Heim2002}. The use
of an erroneous statistical distribution that does not fit the data is
usually the highest source of uncertainty \citep{Laimighofer2022}. In
the case of Chile, due to its high degree of climatic variability, it is
difficult to choose a statistical distribution that can be used across
its entire extent. We therefore use a non-parametric method for the
calculation of the drought indices, following \citep{Farahmand2015}.

@@

We also used the MODIS product (MOD13A3\citep{Didan2015}) to calculate
vegetation productivity, and calculated anomalies of cumulative NDVI
using zcNDVI \citep{Zambrano2018}, which was derived from the monthly
time series of NDVI, with Equations 2 and 4. For vegetation
productivity, we selected the time scale that best correlates with
annual net primary productivity (NPP) across continental Chile. For this
purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months
(from December) and compared it with the annual NPP. We obtained NPP
from MOD17A3HGF90. We chose to use six months because the \(R^2\)
between zcNDVI and NPP reaches its highest value at six months,
obtaining an \(R^2\) of 0.31 for forest and 0.72 for shrubland
(Supplementary Information Section S2). We subsequently used zcNDVI with
a time scale of 6 months and calculated it at a monthly frequency for
2000--2023.

@@

We discovered that the Atacama desert, Central Andean dry puna, and the
Magellanic subpolar forests experience an increase in water supply (SPI,
SSI), as well as an increase in atmospheric and vegetation water demand
(EDDI, SETI). However, in the Magellanic subpolar forests, we found no
evidence of either a significant increase or decrease in SETI across
time scales. Also, we found a significant decreasing trend in water
supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean
Matorral\citep{Boisier2018, Sarricolea2019}, Valdivian temperate
forests, and Patagonian steppe, accompanied by an increase in
atmospheric water demand (EDDI). Our results indicate that temporal
trends of water supply and atmospheric demand tend to decrease or
increase more strongly over longer time scales,a trend that is
consistent with the progressive intensification of drought severity
across much of Chile, and that has been observed in other regions facing
long-term droughts\citep{Rashid2019, Miro2023}. Simultaneously, we
observed a divergent trend between EDDI and SETI. In the majority of
ecoregions, a rise in atmospheric water demand (EDDI) typically leads to
a rise in vegetation water demand (SETI). However, in the ecoregions
most affected by drought (Fig. 3 and Fig. 5), i.e., the Chilean
matorral, and the Patagonian steppe , we found that an increase in
atmospheric water demand results in a decrease in the water demand of
vegetation. Together, our findings demonstrate a persistent drying trend
in the Chilean Matorral, the Patagonian steppe, and the Southern Andean
steppe. We attribute this trend to a simultaneous decrease in
precipitation and an increase in atmospheric evaporative demand, leading
to a decrease in the water demand by vegetation in water-limited
areas\citep{Pascoa2021}.

@@

The consequences of the persistent drying trend for ecosystems
throughout continental Chile are manifold. First, the prolonged
hydrological drought, i.e., precipitation deficit, has reduced
groundwater storage (SSI; ref. \citep{Taucare2024}), leading to a steady
decline in vegetation productivity (zcNDVI) since 2000 across the
Patagonian steppe, the Southern Andean steppe, and the Chilean Matorral,
which reached its lowest level between 2020 and 2022 and could be due to
either a decrease in vegetation area, a loss of biomass, or browning in
forest ecosystems. Recent studies examining natural and productive
ecosystems\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021} have attributed
the decline in vegetation productivity with declines in soil moisture
and increases in evapotranspiration. Second, the sharp decline in
vegetation productivity in the Chilean Matorral and Valdivian temperate
forest ecoregions showed that grasslands and shrublands respond to
shifts in water supply over longer time scales (12 months) than savannas
and croplands (6 months). Also, in the Valdivian temperate forest
ecoregion, which has a large forested area, vegetation productivity
responded to soil moisture (SSI) and vegetation water demand (SETI) most
strongly at 12 and 36 months, respectively. This result is consistent
with recent studies showing that progressive, long-term water deficits
in central Chile have triggered forest browning and declines in native
forest productivity \citep{Miranda2023, Miranda2020, Venegas2022}. While
our analysis do not distinguish between native and planted forests, the
latter of which are considered to be more drought tolerant in central
and southern Chile\citep{Carrasco2022}, we show that forest area
declines more sharply in response to increasing water demand due to
rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI;
refs. \citep{Fajardo2019, Holz2018}), which may have cascading impacts
on multiple facets of forest diversity\citep{Segovia2021, Sabatini2022}.

Moreover, the strengthening of the correlation between vegetation
productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI)
over multiple time scales (up to 36 months) and across land cover types
(Fig. 5) demonstrates the impacts of climate change on the water balance
in Chile. These impacts may extend beyond vegetation productivity, as
reduced soil moisture in central Chile and the western United States has
increased wildfire activity\citep{Holden2018, Gonzalez2018}, which is a
growing concern in Chile and may be further exacerbated by extensive
plantations of highly flammable tree species, e.g., Eucalyptus spp. and
Pinus spp.\citep{Bowman2019}. Lastly, we found that the decline in the
vegetation productivity of croplands is largely due to a decrease in the
water supply and vegetation water demand to a greater extent than to an
increase in atmospheric water demand\citep{Quiring2010}, causing a
decline in water availability. This is consistent with evidence that
more water-intensive crops have replaced less water-intensive crops in
central Chile, leading to an increase in water extraction from rivers or
groundwater \citep{Munoz2020, Duran2020}.

@@

We found evidence that temporal decreases in water supply (SPEI, SSI)
and decreases in vegetation water demand (SETI) are driving shifts not
only in vegetation productivity but also in temporal trends of land
cover change across most of continental Chile. Despite differences in
drought tolerance (e.g., shrublands, grasslands, and savannas), our
results provide evidence that the area of most land cover types
dominated by vegetation has been affected by water deficits, albeit to
varying degrees (Fig. 8). Additionally, our results suggest that water
deficits, to a greater extent than factors associated with human
activity, have affected temporal trends in land cover change for most
land cover types. Further, we found that temporal changes in cropland
cover may not be a direct consequence of drought (Fig. 7), but rather an
indirect one, likely associated with forest fires and possibly the
decisions of resource-poor farmers to migrate to regions with more
abundant water resources or to change economic
activity\citep{AghaKouchak2021, Hermans2021}. The reason for the
non-linear increases in forest area in response to burned area across
most ecoregions (Fig. 8) is unclear, as it could be due to forest
recovery\citep{UrrutiaJalabert2018} or the establishment of forest
plantations \citep{Smith2021}.

@@

Our analysis of the impacts of water supply and demand on vegetation
productivity and land cover change has some important limitations that
need to be highlighted.. One of the principal limitations of this study
is the use of secondary information. For instance, we used estimates of
water supply and demand, such as ERA5L and MODIS, which, despite their
improved estimation capacity, suffer from biases and
uncertainties\citep{Gomis2023, Clelland2024} in different areas or
climatic conditions. In this study, we compared the ERA5L data with
climatic stations (see Table S2) to verify bias and uncertainty, but
future studies that aim for more focalized analysis will need to improve
the precision of these products. We used zcNDVI\citep{Zambrano2018}
(MODIS) as a proxy for vegetation productivity, which has proven to be a
good estimate of NPP (see Fig. S1 and S2), but its quality varies
between different types of vegetation.

A second limitation is that we used products that estimate land cover
types using classification models, which are subject to quality errors
that must be taken into account\citep{Stehman2019, Verburg2011}. In
addition, in our case we used macro classes of land cover, where, for
example, the different types of forests (e.g., monoculture, native
forest) were pooled into the same land cover type. This approach may
hinder our ability to understand the effects of drought on the various
subclasses within each land cover class. In terms of croplands, we could
not distinguish between rainfed and irrigated areas using macro classes.
However, in this study, we aimed to provide a broad overview at a large
spatial scale, but acknowledge that using sub-classes of land cover
types at finer spatial resolutions may help to better understand
underlying mechanisms.

In our analysis of the impacts of drought intensity on temporal trends
of land cover change, we integrated proxies for human activity that also
may affect land cover change. However, attributing land cover change to
human activity and decisions is complex when using earth observation
tools. While earth observation tools can analyze land cover change,
whether a land cover type changes likely depends on a multitude of
social and economic factors that are challenging to
quantify\citep{Rindfuss2004, Jenner2024} and necessitate the integration
of social, natural, and geographic information sciences.

\section{Conclusion}\label{conclusion}

Overall, our results show that long-term declines in water supply and
demand have consistently induced widespread, multi-dimensional impacts
on the vegetation productivity and on the temporal trends of changes in
land cover types across a broad range of ecoregions. While prolonged
droughts may directly cause shifts to more drought-tolerant land cover
types, such as shrublands, we also found that areas affected by fires
were associated with increases in the area of forests and croplands,
highlighting the importance of socio-economic factors in shaping land
use change dynamics. Our study extends current understanding of drought
impacts by demonstrating how their multidimensionality emerges over
multiple time scales and across land cover types, which can contribute
to developing context-specific adaptation strategies for agriculture,
biodiversity conservation, and natural resource management.
